Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-19 index.qmd
section:background,dag; process:write

<!-- 5. continuing with the motivating example -->
The motivating example in @sec-framework-structural can further illustrate how to use the five fundamental building blocks to construct a system's causal structure. In this scenario, the investigator aims to determine whether, *on average*, there is a difference in conceptual understanding of a topic $(Y_{i}=\{0,1\})$ between two groups of students $(T_{i}=\{1,2\})$, described by the estimand $E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$. However, unlike the previous case, an experiment cannot be conducted, and the problem suggests that the country to which a student belongs ($X$) may influence both $T$ and $Y$. Such scenarios are plausible, especially when the teaching methods depend on software or access to technology, which may be limited in certain countries [(maybe an example with more impact?)]{style="color:blue;"}. @fig-example illustrates the plausible causal structure of the motivating example. A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks. The figures display multiple descendants, as indicated by pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. Additionally, the figures features multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Finally, the figures illustrate the fork $X \rightarrow \{T,Y\}$, and two colliders with $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.

>>>>

<!-- 5. continuing with the motivating example -->
The motivating example in @sec-framework-structural can further illustrate how to use the five fundamental building blocks to construct a system's causal structure. In this scenario, the investigator aims to determine whether, *on average*, there is a difference in conceptual understanding of a topic $(Y=\{0,1\})$ between two groups of students $(T=\{1,2\})$, described by the estimand $E[Y| T=1] - E[Y| T=2]$. However, unlike the previous case, an experiment cannot be conducted, and the problem suggests that the country to which a student belongs ($X$) may influence both $T$ and $Y$. Such scenarios are plausible, especially when the teaching methods depend on software or access to technology, which may be limited in certain countries [(maybe an example with more impact?)]{style="color:blue;"}. @fig-example illustrates the plausible causal structure of the motivating example. A detailed examination of Figures [-@fig-mdag_example1], [-@fig-sdag_example1], and SCM [-@fig-scm_example1] reveals the presence of at least four of the five fundamental building blocks. The figures display multiple descendants, as indicated by pairwise relations such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. Additionally, the figures features multiple pairs of unconnected nodes, evident from the relations $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Finally, the figures illustrate the fork $X \rightarrow \{T,Y\}$, and two colliders with $\{X,e_{T}\} \rightarrow T$ and $\{X,T,e_{Y}\} \rightarrow Y$.
